---
title: "The Problem"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Sure, it's easy to identify team roles when observing a single match at a time, but it's not humanly possible to identify team roles for the hundreds, thousands, and millions of matches that have run through Riot Games' servers through eyeballs alone.  

Fortunately, Riot Games provides post-match data of teams and individual players for virtually all matches that go through their servers.  **Is it possible for a machine to take a sample of data from these matches and be able to distinguish all five team roles from each other?**


### Why Solving the Problem Matters

In the Summoner's Rift game modes for Classic 5v5 matches, Draft Pick and Ranked, because the players' selection of champions are done and revealed in a structured order, team roles are always assigned the same Participant IDs according to which side of the map the players are on:

```{r echo = FALSE, results = "asis"}
library(knitr)
TeamRole <- c("Top", "Jungle", "Mid", "BotCarry", "Support")
BlueTeam <- c(1,2,3,4,5)
RedTeam <- c(6,7,8,9,10)
df_roles <- data.frame(TeamRole, BlueTeam, RedTeam)
kable(df_roles, caption = "Figure 1: Assumed Team Role Participant IDs -- Draft Pick, Ranked, eSports")
```

While nothing is preventing teammates from swapping team roles before a match or even during a match (e.g., a ***Top Laner*** picks Smite as a summoner spell or has the Unsealed Spellbook keystone rune and swaps their current summoner spell for Smite and becomes a de facto ***Jungler***), it's safe to assume that the assignment of these team roles are cemented in place.

However, in the other Summoner's Rift game mode, Blind Pick, the selection of all ten champions are done and revealed simultaneously, and, therefore, team roles cannot be assumed to be in the same Participant ID order, nor can it be assumed that the five distinct team roles exist on each side as a match proceeds.  For instance, one team may have two ***Junglers***, while the other team has no ***Junglers***.

Furthermore, while Riot Games appears to have a classification system in place for labeling a player in a certain lane and role, it has shown to be flawed, even in a game mode with an ordered champion selection as is the case in professional eSports matches.  Below is a table representing the NA LCS 2018 Spring Split team role classifications for all 117 matches (thus, 1170 classifications).  The column names are team roles that I manually entered according to Participant ID, while the row names represent a concatentation of the ```role``` and ```lane``` variables that Riot Games assigned and were obtained from the data set:


```{r message = FALSE, results = "asis"}
library(dplyr)
nalcs_season_match_data <- read.csv(
  "../datasets/nalcs/nalcs_spring2018_match_player_stats.csv") %>% 
  mutate(roleLane = paste(role, lane, sep = ", "))
table_roles <- table(nalcs_season_match_data$roleLane, nalcs_season_match_data$teamRole)
kable(table_roles, caption = "NA LCS Spring Split 2018 Team Role Assignments -- Predicted Roles (Rows) vs. Actual Roles (Columns)")
```
```{r echo = FALSE}
remove(nalcs_season_match_data)
```

Not only are there ten different combinations of Role and Lane variables in this data set, but there are a large number of assumed roles being auto-classified into different roles.  

For instance, there are a significant number of ***Mid*** and ***Top Laners*** being assigned the ```NONE, JUNGLE``` ```role-lane``` combination.  This can be theorized in one way: Riot Games has a classification system in place that assigns players to a ```role-lane``` according to where they are located on the map at a certain point in time in a match, particularly early.   

One reason to make this theory is because there are two popular tactics used before the laning phase.  One is **counter-jungling** (invading the enemy ***Jungler***), and the other is **leashing** (providing early assistance to the ***Jungler*** teammate in killing their first jungle monster).  One or more non-Jungler teammates are optional for the former but required for the latter.  Therefore, when the time comes for the system to assign the ```role-lane```, those teammates who are still in the jungle may inadvertantly be assigned the ```NONE, JUNGLE``` ```role-lane``` combination.

While the auto-classification system has shown to be flawed, at least it's safe to assume team roles for Draft Pick, Ranked, and eSports matches due to the structure cited in Figure 1, and many third-party League of Legends analytics websites, such as [op.gg](http://na.op.gg), [champion.gg](https://champion.gg), and [mobalytics.gg](https://mobalytics.gg), appear to make that assumption.  However, these sites miss out on analyzing data from Blind Pick matches, in which team role assumptions cannot be made.  If the auto-classification system can be improved upon, then both Riot Games and these third-party websites can utilize this improved system for their future analytics.